When and Where: Understanding the Patterns of Emergency Service Calls in Toronto

### 2024-09-27 paper/paper.qmd
Added removing NA observations
and updated accordingly

---
title: "When and Where: Understanding the Patterns of Emergency Service Calls in Toronto"
subtitle: "Exploring the Temporal and Spatial Dynamics of Reported Emergency Incidents"
author: 
  - Xinxiang Gao
thanks: "Code and data are available at: https://github.com/xgao28/PICCFSA."
date: today
date-format: long
abstract: "This paper examines ten years of data on crisis service calls in Toronto, focusing on overdose, Person in Crisis, and suicide-related incidents. Our analysis reveals clear temporal patterns, with overdoses peaking at night and mental health crises spiking in the afternoon. Spatially, certain neighborhoods such as Moss Park and Downtown Yonge East consistently report higher numbers of crisis events. These findings highlight the importance of targeted mental health and substance abuse interventions, as well as resource allocation strategies tailored to specific times and locations."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The Persons in Crisis Calls for Service dataset, provided by the Toronto Police Service (TPS), offers a comprehensive view of crisis-related incidents attended by TPS. This dataset includes over 309,000 records from January 1, 2014, to June 26, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose. The dataset captures incidents that required police intervention, excluding those handled by other TPS divisions, such as Parking and Marine. Additionally, the dataset does not contain sensitive information related to individuals in crisis, ensuring that no individual can be identified from the results.

Understanding the nature and patterns of crisis calls is crucial for several reasons. For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, insights into the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.

@@

```{r}
#| include: false
#| warning: false
#| message: false
data <- read.csv("../data/analysis_data/piccfsa.csv")
```

```{r summary_stats}
#| label: tbl-summary
#| echo: false
#| warning: false
#| message: false
summary_stats <- data %>%
  group_by(event_type) %>%
  summarise(
    total_count = sum(n())
  ) %>%
  ungroup()

@@

Over the ten-year period, the dataset reveals significant insights into the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary. Among these, Person in Crisis events are by far the most frequent, comprising more than half of the total incidents.

Overdose: There were a total of 34,097 overdose incidents recorded over the decade. This category includes all cases where an individual required police intervention due to drug or alcohol overdose. The number of overdose incidents highlights the ongoing public health challenge related to substance abuse in the community.

Person in Crisis: With 177,851 incidents, Person in Crisis is the most frequently reported type of event. These incidents involve individuals who are experiencing mental health crises or other forms of distress that necessitate police involvement. The high count underscores the critical need for mental health support and resources within the community.

Suicide-related: A total of 97,155 suicide-related incidents were documented. This includes both attempts and situations where there was a risk of suicide, requiring immediate police response. The prevalence of such incidents emphasizes the importance of suicide prevention efforts and access to mental health services.

@@

```{r}
#| label: fig-hour
#| fig-cap: Event Type Counts Grouped by Hour of the Day
#| echo: false
#| warning: false
#| message: false
df_summarized <- data %>%
  group_by(event_hour, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

# Create the line plot
ggplot(df_summarized, aes(x = event_hour, y = count, color = event_type, group = event_type)) + 
  geom_line() + 
  labs(title = "Event Type Counts Grouped by Hour", 
       x = "Hour", 
       y = "Count", 
       color = "Event Type") +
  theme_minimal()
```

@@

```{r}
#| label: fig-dow
#| fig-cap: Event Type Counts Grouped by Day of the Week
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = event_dow, fill = event_type)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Event Type Counts Grouped by Day of the Week", 
       x = "Day of the Week", 
       y = "Count",
       fill = "Event Type") +
  theme_minimal()
```

@@

```{r}
#| label: fig-month
#| fig-cap: Event Type Counts Grouped by the Month
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = event_month, fill = event_type)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Event Type Counts Grouped by the Month", 
       x = "Month", 
       y = "Count",
       fill = "Event Type") +
  theme_minimal() +
  coord_flip()
```

@@

```{r}
#| label: fig-date
#| fig-cap: Event Type Counts Grouped by Date
#| echo: false
#| warning: false
#| message: false
df_summarized <- data %>%
  group_by(event_date, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

@@

ggplot(df_summarized, aes(x = event_date, y = count, color = event_type, group = event_type)) + 
  geom_smooth() + 
  labs(title = "Event Type Counts Over Time", 
       x = "Date", 
       y = "Count", 
       color = "Event Type") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y",
               limits = as.Date(c(min(df_summarized$event_date), max(df_summarized$event_date)))) +
  theme_minimal()
```

@@

```{r}
#| label: fig-hood
#| fig-cap: Event Counts in Neighbourhood Grouped by Event Type
#| echo: false
#| warning: false
#| message: false
# Step 1: Summarize the data by counting occurrences of each event_type
df_summarized <- data %>%
  group_by(hood_158, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

@@

@@

  # Step 3: Create a bar plot for the top 10 neighborhoods for the current event type
  plot <- ggplot(df_top10, aes(x = reorder(hood_158, count), y = count)) + 
    geom_bar(stat = "identity") + 
    labs(title = event, 
         x = "Neighborhood 158", 
         y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    coord_flip()

@@

kable(neighborhood_table, caption = "A Look-up Table for Hood Number and Hood Name", 
             booktabs = TRUE, align = 'l', label = "tab-hood")
```

@@

The analysis of the Persons in Crisis Calls for Service dataset reveals meaningful patterns in both the temporal and spatial distribution of crisis-related events in Toronto over the past decade. Overdose, Person in Crisis, and Suicide-related incidents exhibit distinct daily, weekly, and seasonal trends. Overdose incidents peak at night, Person in Crisis events are most frequent in the afternoon, and Suicide-related incidents show a steady occurrence during the late afternoon and early evening. Additionally, Thursdays see the highest volume of crisis calls across all types, while weekends have lower incident counts. These temporal patterns provide valuable insight for optimizing resource allocation and tailoring response strategies to times of greatest need.

Spatially, certain neighborhoods, such as Moss Park and Downtown Yonge East, show consistently high rates of crisis incidents, particularly for overdoses. These findings highlight the need for targeted interventions in high-risk areas, which could be related to local socioeconomic factors or population density. Understanding both the temporal and spatial dynamics of crisis events is crucial for improving emergency response efforts and supporting the communities most affected.

Understanding when and where these crisis events are most likely to occur is critical for resource allocation and emergency response planning. By highlighting the neighborhoods and time frames with higher incident rates, this analysis offers actionable insights for policymakers and service providers to improve crisis response strategies, ultimately enhancing public safety and community well-being.

This dataset also contains additional information that could further enrich the analysis, such as the police division of the event, whether an official occurrence was created, and whether a Mental Health Act (MHA) apprehension was made. Future work could explore the impact of these factors on crisis management and resolution. Additionally, while this dataset focuses on three main crisis types, expanding the scope to include more detailed emergency call categories could offer a more comprehensive understanding of Toronto’s broader crisis response landscape, providing deeper insights into the city's emergency response system.

@@

- tidyverse [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.

- ggplot2 [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.

- cowplot [@cowplot]: The package enabled us to combine multiple plots into a single figure, enhancing the presentation of our results.

- lubridate [@lubridate]: This package made it straightforward to work with date and time data, simplifying the process of manipulating and analyzing temporal information.

- knitr [@knitr]: This package was crucial for integrating R code and its output directly into the document, allowing for reproducible research and dynamic report generation.

- janitor [@janitor]: This package provided helpful tools for data cleaning. Its functionalities greatly enhanced the efficiency and accuracy of our data preparation process.

@@

>>>>

---
title: "When and Where: Understanding the Patterns of Emergency Service Calls in Toronto"
subtitle: "Exploring the Temporal and Spatial Dynamics of Reported Emergency Incidents"
author: 
  - Xinxiang Gao
thanks: "Code and data are available at: https://github.com/xgao28/PICCFSA."
date: today
date-format: long
toc: true
abstract: "This paper examines ten years of data on crisis service calls in Toronto, focusing on overdose, Person in Crisis, and suicide-related incidents. Our analysis reveals clear temporal patterns, with overdoses peaking at night and mental health crises spiking in the afternoon. Spatially, certain neighborhoods such as Moss Park and Downtown Yonge East consistently report higher numbers of crisis events. These findings highlight the importance of targeted mental health and substance abuse interventions, as well as resource allocation strategies tailored to specific times and locations."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The Persons in Crisis Calls for Service dataset, provided by the Toronto Police Service (TPS), offers a comprehensive view of crisis-related incidents attended by TPS. This dataset includes over 309,000 records from January 1, 2014, to June 30, 2024, focusing on significant event types such as Suicide, Person in Crisis, and Overdose. The dataset captures incidents that required police intervention, excluding those handled by other TPS divisions, such as Parking and Marine. Additionally, the dataset does not contain sensitive information related to individuals in crisis, ensuring that no individual can be identified from the results.

Understanding the nature and patterns of crisis calls is foundational for several reasons. For public safety, identifying high-activity areas and time periods can help in deploying resources more effectively, ensuring quicker response times and better support for individuals in crisis.In terms of resource allocation, statistical results towards the demand for crisis services can inform budgeting and staffing decisions, ensuring that resources are allocated where they are most needed.For policy development, the findings can guide the development of policies and interventions aimed at reducing the incidence of crises and improving community well-being.

@@

```{r}
#| include: false
#| warning: false
#| message: false
data <- read.csv("../data/analysis_data/piccfsa.csv")

# Factor day of week and month for correct ordering
data$event_dow <- factor(data$event_dow, 
                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

month_levels <- c("January", "February", "March", "April", "May", "June",
                  "July", "August", "September", "October", "November", "December")

data$event_month <- factor(data$event_month, levels = month_levels)
```

```{r summary_stats}
#| label: tbl-summary
#| echo: false
#| warning: false
#| message: false
summary_stats <- data %>%
  group_by(event_type) %>%
  summarise(
    total_count = sum(n())
  ) %>%
  ungroup() %>% 
  rename("Event Type" = event_type, "Total Count" = total_count)

@@

## Overview
Over the ten-year period, the dataset reveals the nature and frequency of three primary types of crisis events: Overdose, Person in Crisis, and Suicide-related, as summarized in @tbl-summary. Among these, Person in Crisis events are by far the most frequent, comprising more than half of the total incidents.

- Overdose: There were a total of 32,494 overdose incidents recorded over the decade. This category includes all cases where an individual required police intervention due to drug or alcohol overdose. The number of overdose incidents highlights the ongoing public health challenge related to substance abuse in the community.

- Person in Crisis: With 172,149 incidents, Person in Crisis is the most frequently reported type of event. These incidents involve individuals who are experiencing mental health crises or other forms of distress that necessitate police involvement. The high count underscores the critical need for mental health support and resources within the community.

- Suicide-related: A total of 92,762 suicide-related incidents were documented. This includes both attempts and situations where there was a risk of suicide, requiring immediate police response. The prevalence of such incidents emphasizes the importance of suicide prevention efforts and access to mental health services.

@@

```{r}
#| label: fig-hour
#| fig-cap: Event Type Counts by Hour of the Day
#| echo: false
#| warning: false
#| message: false
df_summarized <- data %>%
  group_by(event_hour, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

# Create the line plot
ggplot(df_summarized, aes(x = event_hour, y = count, color = event_type, group = event_type)) + 
  geom_line() + 
  labs(title = "Event Type Counts by Hour", 
       x = "Hour", 
       y = "Count", 
       color = "Event Type") +
  theme_minimal()
```

@@

```{r}
#| label: fig-dow
#| fig-cap: Event Type Counts by Day of the Week
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = event_dow, fill = event_type)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Event Type Counts by Day of the Week", 
       x = "Day of the Week", 
       y = "Count",
       fill = "Event Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@@

```{r}
#| label: fig-month
#| fig-cap: Event Type Counts by the Month
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = event_month, fill = event_type)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Event Type Counts by the Month", 
       x = "Month", 
       y = "Count",
       fill = "Event Type") +
  theme_minimal() +
  coord_flip()
```

@@

```{r}
#| label: fig-date
#| fig-cap: Event Type Counts Over Time
#| echo: false
#| warning: false
#| message: false
df_summarized <- data %>%
  group_by(event_date, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

@@

ggplot(df_summarized, aes(x = event_date, y = count, color = event_type, group = event_type)) + 
  geom_smooth() + 
  labs(title = "Event Type Counts Over Time", 
       x = "Date", 
       y = "Count", 
       color = "Event Type") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y",
               limits = as.Date(c(min(df_summarized$event_date), max(df_summarized$event_date)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

```{r}
#| label: fig-hood
#| fig-cap: Event Counts in Neighbourhood by Event Type
#| echo: false
#| warning: false
#| message: false
# Step 1: Summarize the data by counting occurrences of each event_type
df_summarized <- data %>%
  group_by(hood_158, event_type) %>%
  summarise(count = n()) %>%
  ungroup()

@@

# Define colors for each event type
event_colors <- c("Overdose" = "#1B9E77",      # Deep green
                  "Person in Crisis" = "#D95F02",  # Muted orange
                  "Suicide-related" = "#7570B3")   # Slate blue

@@

  # Step 3: Create a bar plot with a consistent color for each event type
  plot <- ggplot(df_top10, aes(x = reorder(hood_158, count), y = count)) + 
    geom_bar(stat = "identity", fill = event_colors[event]) +  # Set a fixed color for the bars
    labs(title = event, 
         x = "Neighborhood 158", 
         y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    coord_flip()

@@

kable(neighborhood_table, caption = "Hood Number and Name Reference", 
             booktabs = TRUE, align = 'l', label = "tab-hood")
```

@@

The analysis of the Persons in Crisis Calls for Service dataset shows notable patterns in the timing and location of crisis-related incidents in Toronto over the past decade. Overdose, Person in Crisis, and Suicide-related events each follow distinct daily, weekly, and seasonal cycles. Overdoses tend to occur most frequently at night, Person in Crisis calls are most common in the afternoon, and Suicide-related incidents maintain a steady rate during late afternoon and early evening. Thursdays experience the highest volume of crisis calls across all types, while weekends typically have fewer incidents. These time-based trends help guide the allocation of resources and the adjustment of response plans to align with periods of greatest demand.

In terms of location, areas like Moss Park and Downtown Yonge East consistently report high numbers of crisis calls, particularly for overdoses. These findings suggest the need for focused interventions in these neighborhoods, potentially influenced by socioeconomic conditions or population density. Recognizing both the temporal and geographic patterns of crisis incidents is key to enhancing emergency response strategies and supporting the areas most impacted.

Knowing when and where these incidents are most likely to happen is vital for planning emergency response efforts and distributing resources effectively. By identifying neighborhoods and time periods with higher incident rates, this analysis offers practical guidance for policymakers and service providers to improve crisis management, ultimately contributing to public safety and community well-being.

The dataset also includes additional information that could deepen the analysis, such as police division details, whether an official occurrence was logged, and whether a Mental Health Act (MHA) apprehension occurred. Future studies could examine how these elements influence crisis resolution. Moreover, although this dataset focuses on three crisis types, incorporating a broader range of emergency call categories could provide a more detailed understanding of Toronto's crisis response landscape.

@@

- `tidyverse` [@tidyverse]: A collection of R packages designed for data science, including dplyr, ggplot2, readr, purrr, and others, which greatly facilitated data manipulation, analysis, and visualization.

- `ggplot2` [@ggplot2]: An implementation of the Grammar of Graphics, which allowed us to create complex and aesthetically pleasing visualizations with ease.

- `cowplot` [@cowplot]: The package enabled us to combine multiple plots into a single figure, enhancing the presentation of our results.

- `lubridate` [@lubridate]: This package facilitates the work with date and time data, simplifying the process of manipulating and analyzing temporal information.

- `knitr` [@knitr]: This package enabled us to perform data demonstration with tables.

- `janitor` [@janitor]: This package provided helpful tools for data cleaning. Its functionalities greatly enhanced the efficiency and accuracy of our data preparation process.

@@

# Appendix

## Data Cleaning

In the data cleaning process, the raw dataset was first read from a CSV file and unnecessary columns, including identifiers and irrelevant variables, were removed. The janitor package was used to clean the column names for consistency. Instances of the value "NSA" in character columns were replaced with NA, and any rows containing NA values were subsequently dropped. To ensure proper ordering in visualizations, the day of the week and month were converted to factors with defined levels. Finally, the cleaned dataset was saved as a new CSV file for further analysis. The cleaned dataset contains over 297,000 observations of 10 variables.
